---
title: "Decision Trees and Cross Validation"
author: "Kim Pham"
date: "11/14/2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## 0. Load Libraries
```{r}
library(rpart)
library(dplyr)
```

## 1. Upload data file

Data file description: This data is from a MOOC and contains the following per-student variables:

certified (yes/no) - Whether or not a student paid for the course  
forum.posts (numeric) - How many forum posts a student made throughout the course  
grade (numeric) - A student's average grade for the course exam  
assignment (numeric) - A student's average grade for the course assignments

```{r}
M1 <- read.csv("Data/MOOC1.csv")
M2 <- read.csv("Data/MOOC2.csv")
```


## 2. Build Classification Tree
```{r}
c.tree <- rpart(certified ~ forum.posts + grade + assignment, method="class", data=M1, control=rpart.control(minsplit=15, cp=.00001))

printcp(c.tree) # print results

post(c.tree, file = "tree.ps", title = "Decision Tree") # plot tree 
#c.tree
```

## 3. Understand Cross Validation Error

The heading "xerror" in the printcp table stands for "cross validation error". It is the error rate of assigning students to certified/uncertified of the model averaged over 10-fold cross validation. 

CP stands for "Cost Complexity" and represents the cost in error for adding a node to the tree. Notice it decreases as we add more nodes to the tree which implies that more nodes make better predictions. However, more nodes also mean that we may be making the model less generalizable, this is known as "overfitting".

If we are worried about overfitting we can remove nodes form our tree using the prune() command, setting cp to the CP value from the table that corresponds to the number of nodes we want the tree to terminate at. Let's set it to two nodes.

## 4. Pruning Tree
```{r}
c.tree2 <- prune(c.tree, cp = 0.00245098)
printcp(c.tree2)
post(c.tree2, file = "tree2.ps", title = "MOOC") 
```

Tree2 has a lower error rate because there are less leaves.

## 5. Use trees to predict student certification on M2 data set.

```{r}
M2$predict1 <- predict(c.tree, M2, type = "class")
M2$predict2 <- predict(c.tree2, M2, type = "class")

t1 <- table(M2$certified, M2$predict1)
t1
t2 <- table(M2$certified, M2$predict2)
t2
```

```{r}
#Calculating Accuracy for T1 and T2 
sum(diag(t1)/sum(t1))
sum(diag(t2)/sum(t2))
```

