---
title: "K-Means Clustering: Home Regions"
author: "Kim Pham"
date: "10/27/2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## 0. Load libraries
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```

## 1. Upload data file

Data file description: This file contains survey answers to the questions and regions where people grew up. 

Code Book:

Duration (in seconds)
Q1 - First Name  
Q2 - Last Name  
Q3 - Have you ever owned a cat?  
Q4 - Do you pronounce "gif", with a J (j-iff) or a G (g-iff)?  
Q5 - How many months have you lived in New York City?  
Q6 - How many siblings (brothers/sisters) do you have?  
Q7 - How many times do you play sport each week?  
Q8 - How many miles do you travel from home to TC?  
Q9 - Estimate how many of your friends own Android phones  
Q10 - How many movies have you seen in the cinema this year?  
Q11 - How many classes are you taking this semester?  
Q12 - How many states have you visited in the US?  
Q13 - What city/town did you grow up in?  
Q14 - What state/province did you grow up in?  
Q15 - What country did you grow up in?  

```{r}
df <- read.csv("Data/HUDK4050_2017_Cluster_Survey.csv")
```

## 2. Process and structure data for kmeans

```{r}
df <- select(df, -first, -last, -city.grew, -state.grew)
df_answers <- select(df, -country.grew, -cat, -gif.jiff) # remove text columns
#df_answers$cat <- as.integer(df_answers$cat) # convert to integer 
#df_answers$gif.jiff <- as.integer(df_answers$gif.jiff) # convert to integer
df_answers[!complete.cases(df_answers),] # no NAs exist
df_s <- scale(df_answers) # scale data
head(df_s)
summary(df_s)
```

## 3. Run K-means using 2 clusters

```{r}
fit <- kmeans(df_s, 2, nstart=50, iter.max=15)
df_cluster <- data.frame(df, fit$cluster)
head(df_cluster)
```

## 4. Visualize clusters

## Prepare data
```{r}
df_cluster$country.grew <- gsub(" ", "", df_cluster$country.grew)
df_cluster$country.grew <- gsub("UnitedStates", "United States", df_cluster$country.grew)
```

## Plot Country Grew Up In and Months in NYC
```{r}
ggplot(df_cluster, aes(country.grew, live.nyc, colour=as.factor(fit.cluster))) + geom_point() + xlab("Country Grew Up In") + ylab("Months in NYC")
```

## Scatterplot Matrix
```{r}
# Basic Scatterplot Matrix
pairs(df_cluster[,3:6], col=df_cluster$fit.cluster)
pairs(df_cluster[,7:10], col=df_cluster$fit.cluster)
```

## 5. Analyze

```{r}
count(df_cluster, fit.cluster)
```

Observations in cluster 2 have grown up mostly in China, Indoneisa, and Taiwan, while observations in cluster 1 have grown up mostly in the United States.

The scatterplot matrix shows relationships between the different variables grouped by clusters. The extreme differences seem to be the number of states traveled to; the number of classes taken; and the number of siblings. 

### Determine number of clusters using elbow plot

We should consider using the elbow plot to determine optimal number of clusters to use. The elbow plot looks at the total within-clusters sum of squares that results from adding a cluster to the kmeans k, or clustering size.

```{r}
# Compute and plot wss for k = 2 to k = 15.
k.max <- 15
wss <- sapply(1:k.max, 
              function(k){kmeans(df_s, k, nstart=50,iter.max = 15 )$tot.withinss})
wss
plot(1:k.max, wss,
     type="b", pch = 19, frame = FALSE, 
     xlab="Number of clusters K",
     ylab="Total within-clusters sum of squares")
```

The elbow plot shows optimal number of clusters K is around 6 or 7 for this dataset. However since the n size is low, the recommendation then might be to revisit the scaled variables the kmeans is running on and rerun the elbow plot.