---
title: "Final Project: K-Means on Acceleration Data"
author: "Kim Pham"
date: "12/12/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

Project: We will apply the K-means clustering algorithm to the acceleration data we collected about Wii Bowling Arm Swings using the Sensor Kinetics App.

## 0. Load libraries
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```

## 1. Upload data file

## 2. Format data to fit kmeans 

We need to combine the aggregated data from all 5 participants into one master dataframe. 

```{r}
setwd("~/Desktop/HUDK4050/assignment-8/Assignment_8_Data/Clean_Data")
file.names <- list.files()
DY_data <- read.csv("dy-clean data with scores.csv")
KP_data <- read.csv("KPmaster.csv")
NA_data <- read.csv("NA_master.csv")
TZ_data <- read.csv("tz_acc_combined.csv")
ZY_data <- read.csv("ZY_master.csv")
all_data <- rbind(DY_data, KP_data, NA_data, TZ_data, ZY_data)
str(all_data)
summary(all_data)
```

"The algorithm will treat each row as a value belonging to a person" so we need to make each row descriptive of one trial. We will do this by grouping by name, filtering by each trial attempt (1-10), and summarizing the mean of the XYZ coordinates of the acceleration.

As a result, we should total 50 rows to equal the 10 attempts of 5 participants.

Note: We have also taken the absolute value of each accleration prior to calculating the mean.

```{r}
x_1 <- all_data %>% group_by(Name) %>% filter(Trial == 1) %>% summarise(avg_X_acc = mean(abs(acc_X_value)))
y_1 <- all_data %>% group_by(Name) %>% filter(Trial == 1) %>% summarise(avg_Y_acc = mean(abs(acc_Y_value))) %>% select(-Name)
z_1 <- all_data %>% group_by(Name) %>% filter(Trial == 1) %>% summarise(avg_Z_acc = mean(abs(acc_Z_value))) %>% select(-Name)
sum_data1 <- bind_cols(x_1, y_1, z_1)
sum_data1$Trial <- 1
Score <- all_data %>% group_by(Name) %>% filter(Trial == 1) %>% summarise(Score = mean(Score)) %>% select(-Name)
sum_data1 <- bind_cols(sum_data1, Score)
```

```{r}
x_2 <- all_data %>% group_by(Name) %>% filter(Trial == 2) %>% summarise(avg_X_acc = mean(abs(acc_X_value))) 
y_2 <- all_data %>% group_by(Name) %>% filter(Trial == 2) %>% summarise(avg_Y_acc = mean(abs(acc_Y_value))) %>% select(-Name)
z_2 <- all_data %>% group_by(Name) %>% filter(Trial == 2) %>% summarise(avg_Z_acc = mean(abs(acc_Z_value))) %>% select(-Name)
sum_data2 <- bind_cols(x_2, y_2, z_2)
sum_data2$Trial <- 2
score <- all_data %>% group_by(Name) %>% filter(Trial == 2) %>% summarise(Score = mean(Score)) %>% select(-Name)
sum_data2 <- bind_cols(sum_data2, Score)
```

```{r}
x_3 <- all_data %>% group_by(Name) %>% filter(Trial == 3) %>% summarise(avg_X_acc = mean(abs(acc_X_value))) 
y_3 <- all_data %>% group_by(Name) %>% filter(Trial == 3) %>% summarise(avg_Y_acc = mean(abs(acc_Y_value))) %>% select(-Name)
z_3 <- all_data %>% group_by(Name) %>% filter(Trial == 3) %>% summarise(avg_Z_acc = mean(abs(acc_Z_value))) %>% select(-Name)
sum_data3 <- bind_cols(x_3, y_3, z_3)
sum_data3$Trial <- 3
score <- all_data %>% group_by(Name) %>% filter(Trial == 3) %>% summarise(Score = mean(Score)) %>% select(-Name)
sum_data3 <- bind_cols(sum_data3, Score)
```

```{r}
x_4 <- all_data %>% group_by(Name) %>% filter(Trial == 4) %>% summarise(avg_X_acc = mean(abs(acc_X_value))) 
y_4 <- all_data %>% group_by(Name) %>% filter(Trial == 4) %>% summarise(avg_Y_acc = mean(abs(acc_Y_value))) %>% select(-Name)
z_4 <- all_data %>% group_by(Name) %>% filter(Trial == 4) %>% summarise(avg_Z_acc = mean(abs(acc_Z_value))) %>% select(-Name)
sum_data4 <- bind_cols(x_4, y_4, z_4)
sum_data4$Trial <- 4
score <- all_data %>% group_by(Name) %>% filter(Trial == 4) %>% summarise(Score = mean(Score)) %>% select(-Name)
sum_data4 <- bind_cols(sum_data4, Score)
```

```{r}
x_5 <- all_data %>% group_by(Name) %>% filter(Trial == 5) %>% summarise(avg_X_acc = mean(abs(acc_X_value)))
y_5 <- all_data %>% group_by(Name) %>% filter(Trial == 5) %>% summarise(avg_Y_acc = mean(abs(acc_Y_value))) %>% select(-Name)
z_5 <- all_data %>% group_by(Name) %>% filter(Trial == 5) %>% summarise(avg_Z_acc = mean(abs(acc_Z_value))) %>% select(-Name)
sum_data5 <- bind_cols(x_5, y_5, z_5)
sum_data5$Trial <- 5
score <- all_data %>% group_by(Name) %>% filter(Trial == 5) %>% summarise(Score = mean(Score)) %>% select(-Name)
sum_data5 <- bind_cols(sum_data5, Score)
```

```{r}
x_6 <- all_data %>% group_by(Name) %>% filter(Trial == 6) %>% summarise(avg_X_acc = mean(abs(acc_X_value)))
y_6 <- all_data %>% group_by(Name) %>% filter(Trial == 6) %>% summarise(avg_Y_acc = mean(abs(acc_Y_value))) %>% select(-Name)
z_6 <- all_data %>% group_by(Name) %>% filter(Trial == 6) %>% summarise(avg_Z_acc = mean(abs(acc_Z_value))) %>% select(-Name)
sum_data6 <- bind_cols(x_6, y_6, z_6)
sum_data6$Trial <- 6
score <- all_data %>% group_by(Name) %>% filter(Trial == 6) %>% summarise(Score = mean(Score)) %>% select(-Name)
sum_data6 <- bind_cols(sum_data6, Score)
```

```{r}
x_7 <- all_data %>% group_by(Name) %>% filter(Trial == 7) %>% summarise(avg_X_acc = mean(abs(acc_X_value)))
y_7 <- all_data %>% group_by(Name) %>% filter(Trial == 7) %>% summarise(avg_Y_acc = mean(abs(acc_Y_value))) %>% select(-Name)
z_7 <- all_data %>% group_by(Name) %>% filter(Trial == 7) %>% summarise(avg_Z_acc = mean(abs(acc_Z_value))) %>% select(-Name)
sum_data7 <- bind_cols(x_7, y_7, z_7)
sum_data7$Trial <- 7
score <- all_data %>% group_by(Name) %>% filter(Trial == 7) %>% summarise(Score = mean(Score)) %>% select(-Name)
sum_data7 <- bind_cols(sum_data7, Score)
```

```{r}
x_8 <- all_data %>% group_by(Name) %>% filter(Trial == 8) %>% summarise(avg_X_acc = mean(abs(acc_X_value)))
y_8 <- all_data %>% group_by(Name) %>% filter(Trial == 8) %>% summarise(avg_Y_acc = mean(abs(acc_Y_value))) %>% select(-Name)
z_8 <- all_data %>% group_by(Name) %>% filter(Trial == 8) %>% summarise(avg_Z_acc = mean(abs(acc_Z_value))) %>% select(-Name)
sum_data8 <- bind_cols(x_8, y_8, z_8)
sum_data8$Trial <- 8
score <- all_data %>% group_by(Name) %>% filter(Trial == 8) %>% summarise(Score = mean(Score)) %>% select(-Name)
sum_data8 <- bind_cols(sum_data8, Score)
```

```{r}
x_9 <- all_data %>% group_by(Name) %>% filter(Trial == 9) %>% summarise(avg_X_acc = mean(abs(acc_X_value)))
y_9 <- all_data %>% group_by(Name) %>% filter(Trial == 9) %>% summarise(avg_Y_acc = mean(abs(acc_Y_value))) %>% select(-Name)
z_9 <- all_data %>% group_by(Name) %>% filter(Trial == 9) %>% summarise(avg_Z_acc = mean(abs(acc_Z_value))) %>% select(-Name)
sum_data9 <- bind_cols(x_9, y_9, z_9)
sum_data9$Trial <- 9
score <- all_data %>% group_by(Name) %>% filter(Trial == 9) %>% summarise(Score = mean(Score)) %>% select(-Name)
sum_data9 <- bind_cols(sum_data9, Score)
```

```{r}
x_10 <- all_data %>% group_by(Name) %>% filter(Trial == 10) %>% summarise(avg_X_acc = mean(abs(acc_X_value)))
y_10 <- all_data %>% group_by(Name) %>% filter(Trial == 10) %>% summarise(avg_Y_acc = mean(abs(acc_Y_value))) %>% select(-Name)
z_10 <- all_data %>% group_by(Name) %>% filter(Trial == 10) %>% summarise(avg_Z_acc = mean(abs(acc_Z_value))) %>% select(-Name)
sum_data10 <- bind_cols(x_10, y_10, z_10) 
sum_data10$Trial <- 10
score <- all_data %>% group_by(Name) %>% filter(Trial == 10) %>% summarise(Score = mean(Score)) %>% select(-Name)
sum_data10 <- bind_cols(sum_data10, Score)
```

Finally, we combine all dataframes into one dataframe, sum_data.

```{r}
sum_data <- bind_rows(sum_data1, sum_data2, sum_data3, sum_data4, sum_data5, sum_data6, sum_data7, sum_data8, sum_data9, sum_data10)
```

## 3. Run kmeans 

We will create clusters using the XYZ accleration combined data and interpret the clusters by observing the clusters along a X-Y plot of score and acceleration XYZ.

####Scale the acceleration XYZ data.

```{r}
scale_acc <- select(sum_data, avg_X_acc, avg_Y_acc, avg_Z_acc)
scale_acc <- scale(scale_acc)
scale_acc <- as.data.frame(scale_acc)
```

```{r}
sum_data <- select(sum_data, -avg_X_acc, -avg_Y_acc, -avg_Z_acc)
sum_data <- bind_cols(sum_data,scale_acc)
```

####Run kmeans with k=3 clusters.

```{r}
fit <- kmeans(scale_acc, 3)
sum_data$cluster <- fit$cluster
```

## 4. Visualize

```{r}
sum_data$cluster <- as.factor(sum_data$cluster)
```

Visualization for avg XYZ accleration per trial by each participant. 

```{r}
ggplot(sum_data, aes(Trial, avg_X_acc, colour = Name)) + geom_line(stat="identity") + xlab("Trial") + ylab("Average X Acceleration") + ggtitle("Avg X Acceleration per Trial by Each Participant") + theme(plot.title = element_text(hjust = 0.5))

ggplot(sum_data, aes(Trial, avg_Y_acc, colour = Name)) + geom_line(stat="identity") + xlab("Trial") + ylab("Average Y Acceleration") + ggtitle("Avg Y Acceleration per Trial by Each Participant") + theme(plot.title = element_text(hjust = 0.5))

ggplot(sum_data, aes(Trial, avg_Z_acc, colour = Name)) + geom_line(stat="identity") + xlab("Trial") + ylab("Average Z Acceleration") + ggtitle("Avg Z Acceleration per Trial by Each Participant") + theme(plot.title = element_text(hjust = 0.5))
```

Visualization for avg XYZ acceleration compared to scoreby clusters.

Note: We have removed observed outliers for X and Y accelerations to improve visualization of the clusters. 

```{r}
x <- filter(sum_data, avg_X_acc < 3)
y <- filter(sum_data, avg_Y_acc > -1.75)

ggplot(x, aes(avg_X_acc, Score, colour = cluster)) + geom_jitter(stat="identity") + xlab("Average X Acceleration") + ylab("Score") + geom_smooth(method = 'lm', se = FALSE) + ggtitle("Clusters by Score and Average X Acceleration") + theme(plot.title = element_text(hjust = 0.5))

ggplot(y, aes(avg_Y_acc, Score, colour = cluster)) + geom_jitter(stat="identity") + xlab("Average Y Acceleration") + ylab("Score") + geom_smooth(method = 'lm', se = FALSE) + ggtitle("Clusters by Score and Average Y Acceleration") + theme(plot.title = element_text(hjust = 0.5))

ggplot(sum_data, aes(avg_Z_acc, Score, colour = cluster)) + geom_jitter(stat="identity") + xlab("Average Z Acceleration") + ylab("Score") + geom_smooth(method = 'lm', se = FALSE) + ggtitle("Clusters by Score and Average Z Acceleration") + theme(plot.title = element_text(hjust = 0.5))
```

## 5. Interpret clusters

From the visualizations, we see the following:

Cluster 1 - positive x, y, z acceleration, AND scores > 6.
Cluster 2 - negative x, y, positive z, AND scores > 6 and scores < 8.
Cluster 3 - positive and negative x (range from -2,2), y ~= 0, negative z, AND scores > 9 and scores < 4 (with 1 outlier score around 7).