---
title: "Principle Component Analysis"
author: "Kim Pham"
date: "11/10/2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## 0. Load libraries
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(corrplot)
library(corrgram)
```

## 1. Upload data file

Data file description: The data comes from the Assistments online intelligent tutoring system (https://www.assistments.org/). It describes students working through online math problems. Each student has the following data associated with them:

- id
- prior_prob_count: How many problems a student has answered in the system prior to this session
- prior_percent_correct: The percentage of problems a student has answered correctly prior to this session
- problems_attempted: The number of problems the student has attempted in the current session
- mean_correct: The average number of correct answers a student made on their first attempt at problems in the current session
- mean_hint: The average number of hints a student asked for in the current session
- mean_attempt: The average number of attempts a student took to answer a problem in the current session
- mean_confidence: The average confidence each student has in their ability to answer the problems in the current session

```{r}
df <- read.csv("Data/Assistments-confidence.csv")
df <- select(df, -id) # remove id
```

## 2. Explore coorelation between variables

```{r}
COR <- cor(df) #coorelation matrix

# visualize coorelation matrix
corrplot(COR, order="AOE", method="circle", tl.pos="lt", type="upper",        
tl.col="black", tl.cex=0.6, tl.srt=45, 
        addCoef.col="black", addCoefasPercent = TRUE,
        sig.level=0.50, insig = "blank")
```

## 3. Process and structure data for PCA

```{r}
df2 <- select(df, -mean_correct) # remove mean_correct variable
df3 <- scale(df2, center = TRUE) # scale data
```

## 4. Run PCA on scaled data

```{r}
pca <- prcomp(df3, scale = TRUE)
```

princomp does not generate eigenvalues directly, but we can print a list of the standard deviation of the variance accounted for by each component.

```{r}
pca$sdev
pca$sdev^2

summary(pca) # pca summary gives the proportion of variance accounted for by each component
plot(pca, type = "lines")
```

## 5. Analyze: 

### Drop PCA components

We should consider dropping components 4-6 since they have the least amount of variances. The limitation is sometimes dropping components result in losing information. For example, if we drop PC4, we could lose useful information, since PC4 has some positive correlation to the mean_correct variable.

```{r}
df_pca <- as.data.frame(pca$x)
df_pca <- cbind(df_pca, as.data.frame(df$mean_correct))
COR2 <- cor(df_pca)
corrplot(COR2)
```
### Examine eigenvectors, or loadings

```{r}
pca$rotation # Eigenvalue = amount of variance/direction ... note: the bigger the eigenvalue, the better. e.g. if it’s 0, the transformation component is not good. It was bad. Maybe the data was hard to characterize.
loadings <- abs(pca$rotation) # abs() will make all eigenvectors positive
sweep(loadings, 2, colSums(loadings), "/") #sweep() computes each row as a proportion of the column.
biplot(pca) # plots data transformation by PC1 and PC2, where axes represent direction of maximum variance and original variable directions are mapped as red arrows; meant to answer which variables "go together" and represent the same underlying construct point in the same direction 
```

* PC1: high variance for mean_hint, mean_atempt, and problems_attempted
* PC2: high variance for prior_percent_correct
* PC3: high variance for mean_confidence
* PC4: high variance for prior_prob_count
* PC5: high variance for mean_attempt and problems_attempted 
* PC6: high variance for mean_hint

### Correlogram relationship of PC1 / PCA2 to mean_correct
```{r}
corrgram(df, order=TRUE,lower.panel=panel.shade,upper.panel=panel.pts,main="Assistments Online Behavior in PC2/PC1 Order")
```

## PCA on new dataset

## 1. Upload data file

Data file description: Data set from Rod Martin, Patricia Puhlik-Doris, Gwen Larsen, Jeanette Gray, Kelly Weir at the University of Western Ontario about people's sense of humor.

```{r}
humor <- read.csv("Data/humor_data.csv")
```

## 2. Process and structure data for PCA

```{r}
humor <- scale(humor, center = TRUE)
```

## 3. Run PCA on scaled data

```{r}
pca <- prcomp(humor, scale = TRUE)
```

## 4. Analyze

### Drop PCA components

We should drop keep PCA components 1-5.

```{r}
pca$sdev
pca$sdev^2
summary(pca)
plot(pca, type = "lines")
```
### Examine eigenvectors, or loadings

```{r}
pca$rotation
loadings <- abs(pca$rotation[,1:5])
sweep(loadings, 2, colSums(loadings))
biplot(pca)
```

Hard to read biplot and the loadings between the components are very similar. We should investigate other methods to interpret the components and its relationship to the original data. 